Type-Safe Modular Parsing

### 2017-05-11 SLE_17/paper.tex
Code & content
Some changes.

\usepackage{listings}
\usepackage{syntax}
\usepackage[T1]{fontenc}
\usepackage[scaled=0.85]{beramono}
\usepackage{multirow}

\lstset{
  language=Scala,                % choose the language of the code
  columns=flexible,
  lineskip=-1pt,
  basicstyle=\ttfamily\small,       % the size of the fonts that are used for the code
  numbers=none,                   % where to put the line-numbers
  numberstyle=\ttfamily\tiny,      % the size of the fonts that are used for the line-numbers
  stepnumber=1,                   % the step between two line-numbers. If it's 1 each line will be numbered
  numbersep=5pt,                  % how far the line-numbers are from the code
  backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  showtabs=false,                 % show tabs within strings adding particular underscores
  morekeywords={var, lazy},
  %  frame=single,                   % adds a frame around the code
  tabsize=2,                  % sets default tabsize to 2 spaces
  captionpos=none,                   % sets the caption-position to bottom
  breaklines=true,                % sets automatic line breaking
  breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
  title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
  escapeinside={(**}{**)},          % if you want to add a comment within your code
  keywordstyle=\ttfamily\bfseries,
}

>>>>

\usepackage{listings}
\usepackage{syntax}
\usepackage[T1]{fontenc}
\usepackage[scaled=0.85]{beramono}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{tabularx}

\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,yield,with,val},
  otherkeywords={=>,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

\lstset{
  language=JavaScala,                % choose the language of the code
  columns=flexible,
  lineskip=-1pt,
  basicstyle=\ttfamily\small,       % the size of the fonts that are used for the code
  numbers=none,                   % where to put the line-numbers
  numberstyle=\ttfamily\tiny,      % the size of the fonts that are used for the line-numbers
  stepnumber=1,                   % the step between two line-numbers. If it's 1 each line will be numbered
  numbersep=5pt,                  % how far the line-numbers are from the code
  backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  showtabs=false,                 % show tabs within strings adding particular underscores
  morekeywords={var, lazy},
  %  frame=single,                   % adds a frame around the code
  tabsize=2,                  % sets default tabsize to 2 spaces
  captionpos=none,                   % sets the caption-position to bottom
  breaklines=true,                % sets automatic line breaking
  breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
  title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
  escapeinside={(**}{**)},          % if you want to add a comment within your code
  keywordstyle=\ttfamily\bfseries,
}

### 2017-05-11 SLE_17/sections/Discussion.tex
Some changes.

Figure~\ref{fig:multi} illustrates the corresponding Scala code
that extends the Object Algebra interface, pretty-printing operation and parser. We use two type parameters \inlinecode{E} and \inlinecode{T} for expressions and types. They guarantee that invalid terms such as \inlinecode{int + int} will be rejected by the parser.
Besides lexing, the trait \inlinecode{TypedLamOAParser} also introduces parsers for types, and the new case for expressions.
We use \inlinecode{pTypedLamT} and \inlinecode{pTypedLamE} as copies of current \inlinecode{pT} and \inlinecode{pE}, due to the issue
with \inlinecode{super} in Scala (see discussion in Section~\ref{subsec:parsingwithoa}). \inlinecode{pT} and \inlinecode{pE} are the parsers used for performing recursion.

@@

From the code, we observe that the multi-sorts of Object Algebra interface
express multiple syntax. The isolation for different syntax
is guaranteed by the type system. The following illustrates some client code
for the new parser:

\lstinputlisting[linerange=40-43]{code/src/papercode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
The code works as expected to parse and pretty-print the result.

@@

As many syntactically extensible parsers, our approach also supports
modifying part of existing parsers, but in a type-safe way. We can
easily update existing parsing rules with new implementations, or
eliminate them in extended parsers. This can be useful in many
situations, including when conflicts or ambiguities arise when
composing languages.
As an illustration, suppose we have an untyped lambda abstraction case in a base parser as below. \inlinecode{pLam} parses a lambda symbol, an identifier, a dot and an expression in sequence. Some irrelevant details are omitted.

\lstinputlisting[linerange=76-82]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=BASEPARSER_UNTYPEDLAM

Then we want to replace the untyped lambda abstractions by typed
lambdas. With inheritance and method overriding, it is easy to only
change the implementation of \lstinline{pLam} in the extended parser.
The new \inlinecode{pLam} parses a colon and a type in addition,
between the identifier and the dot. Due to dynamic dispatch, our new
implementation of lambdas will be different without affecting the other parts of the parser.

@@

\begin{itemize}[leftmargin=*]
    \item \textbf{Object Algebra interface:} defined as a trait for the abstract syntax. The type parameters represent multiple sorts of syntax, and the methods are constructs.
    \item \textbf{Parser:} corresponding parser of the abstract syntax, written in a modular way as we demonstrated in previous sections.
    \item \textbf{Object Algebras (optional):} concrete operations on ASTs, such as pretty-printing.
\end{itemize}

@@

Under our modular parsing framework, we use Packrat parsing as the underlying parsing technique, OO inheritance with method overriding for composing and extending parsers, and Object Algebras for parsing extensible ASTs. However, our framework is more general and thus more powerful, because those aspects are orthogonal to each other, and hence can have alternatives.

@@

\item {\bf Parsing Technique:}
Our approach does not depend on a particular parser combinator library
or parsing algorithm. We demonstrated that Packrat parsing has some
advantages and is suitable in a modular setting. However, under
certain circumstances or in other programming languages, it may not be
the best choice. One can use other parser combinator libraries as long
as the guidelines proposed in Section~\ref{sec:packrat} are met.

\item {\bf Composing and Extending Parsers:}
  We use traits to model parsers, OO inheritance to compose them, and
  method overriding for their extensibility. These language features
  may not be supported in other programming languages, especially in
  functional languages such as Haskell. Nevertheless, \textit{open
    recursion}~\cite{CookThesis} could be used as an alternative. An explicit
  ``self-reference'' parameter is able to explain the recursive calls
  dynamically, by the real argument passed at runtime.

\item {\bf Extensible ASTs:}
Besides Object Algebras, many other techniques including \textit{Data
  types à la carte} (DTC)~\cite{swierstra2008data} and the Cake pattern~\cite{odersky2005independently} also support extensible data structures. These alternatives could also be adopted to build ASTs and corresponding parsers.

@@

As we demonstrated, the whole modular parsing framework itself is
modular, which means it can be customized easily.

>>>>

Figure~\ref{fig:multi} illustrates the corresponding Scala code
that extends the Object Algebra interface, pretty-printing operation and parser. We use two type parameters \inlinecode{E} and \inlinecode{T} for expressions and types. They guarantee that invalid terms such as \inlinecode{int + int} will be rejected by the parser.
Besides lexing, the trait \inlinecode{LamOAParser} also introduces parsers for types, and the new case for expressions.
We use \inlinecode{pTypedLamT} and \inlinecode{pTypedLamE} as copies of current \inlinecode{pT} and \inlinecode{pE}, due to the issue
with \inlinecode{super} in Scala (see discussion in Section~\ref{subsec:parsingwithoa}). \inlinecode{pT} and \inlinecode{pE} are the parsers used for performing recursion.

@@

From the code, we observe that the multi-sorts of Object Algebra interface
express multiple syntax. The isolation for different syntax
is guaranteed by the type system. The following client code

\lstinputlisting[linerange=40-43]{code/src/papercode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
works as expected to parse and pretty-print the result.

@@

As many syntactically extensible parsers, our approach also supports
modifying part of existing parsers, including updating or eliminating existing rules,
but in a type-safe way. This can be useful in many
situations, for instance when conflicts or ambiguities arise upon composing languages.
As an illustration, suppose we have an untyped lambda abstraction case in a base parser as below. Some irrelevant details are omitted.

\lstinputlisting[linerange=76-82]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=BASEPARSER_UNTYPEDLAM
Here \inlinecode{pLam} parses a lambda symbol, an identifier, a dot and an expression in sequence.
Then we want to replace the untyped lambda abstractions by typed
lambdas. With inheritance and method overriding, it is easy to only
change the implementation of \lstinline{pLam} in the extended parser.
Due to dynamic dispatch, our new
implementation of lambdas will be different without affecting the other parts of the parser.

@@

\begin{itemize}[leftmargin=*]
    \item \textbf{Object Algebra interface:} defined as a trait for the abstract syntax. The type parameters represent multiple sorts of syntax, and  methods are constructs.
    \item \textbf{Parser:} corresponding parser of the abstract syntax, written in a modular way as we demonstrated before.
    \item \textbf{Object Algebras (optional):} concrete operations on ASTs, such as pretty-printing.
\end{itemize}

@@

Under our modular parsing framework, we use Packrat parsing as the underlying parsing technique, OO inheritance with method overriding for composing and extending parsers, and Object Algebras for parsing extensible ASTs. However, such a pattern is more general and thus more powerful, because those aspects are orthogonal to each other, and hence can have alternatives.

@@

\item {\bf Parsing Technique:}
The pattern does not depend on a particular parser combinator library
or parsing algorithm. We demonstrated that Packrat parsing has some
advantages and is suitable in a modular setting. However, under
other circumstances it may not be
the best choice. One can use other parser combinators as long
as the guidelines proposed in Section~\ref{sec:packrat} are met.

\item {\bf Composing and Extending Parsers:}
  We use traits to model parsers, OO inheritance to compose them, and
  method overriding for their extensibility. These features
  may not be supported in other programming languages, especially in
  functional languages such as Haskell. Nevertheless, \textit{open
    recursion}~\cite{CookThesis} could be used as an alternative. An explicit
  ``self-reference'' parameter is able to explain the recursive calls
  dynamically, by the real argument passed at runtime.

\item {\bf Extensible ASTs:}
Besides Object Algebras, many other techniques including \textit{Data
  types à la carte} (DTC)~\cite{swierstra2008data} and the Cake pattern~\cite{odersky2005independently} also support extensible data structures. These alternatives could also be adopted to build ASTs and their parsers.

@@

As we demonstrated, such a pattern for modular parsing is 
modular itself, and can be customized easily.

### 2017-05-11 SLE_17/sections/Introduction.tex
Some changes.

\paragraph{Algorithmic Challenges} A first challenge was to do with
the parsing algorithms themselves. Since many parsing formalisms
were not designed with extensibility in mind, some of the employed
techniques were non-modular, or performed poorly in a modular
setting.  The most widely use tools for parsing are parser
generators, but most algorithms employed by parser generators
require full information about the grammar to generate parsing
code. Moreover, actions associated with grammar productions are
typically only type-checked after the parser has been
generated. Both problems go against our goals of semantic
modularity.

@@

To address such algorithmic challenges, we propose a methodology
for implementing modular parsers built on top of an
existing Packrat~\cite{Ford2002} parsing library for Scala~\cite{odersky2004overview}. The Packrat parsing library
directly supports left-recursion, and a \emph{longest-match
  composition} operator. The direct support for left-recursion
avoids the need for left-recursion elimination, which is very
cumbersome is a modular setting. The longest-match composition
operator avoids the use of priorities, and it provides reasonably
efficient optimizations for backtracking.

@@

All the code in the paper and the case study is written in Scala. But,
other languages that support some form of multiple inheritance
(including Java 8 with default methods~\cite{Goetz2012}) could in principle be
used. Our choice for Scala had to do with the concise and elegant syntax
of the language which enables a clear presentation of our ideas.

>>>>

\paragraph{Algorithmic Challenges} A first challenge was to do with
the parsing algorithms themselves, since they 
were usually not designed with extensibility in mind. The most widely used tools for parsing are parser
generators, but they mostly require full information about the grammar to generate parsing
code. Moreover, actions associated with grammar productions are
typically only type-checked after the parser has been
generated. Both problems go against our goals of semantic
modularity.

@@

To address such algorithmic challenges, we propose a methodology
for implementing modular parsers built on top of an
existing Packrat~\cite{Ford2002} parsing library for Scala~\cite{odersky2004overview}. Such a library
directly supports left-recursion, memoization, and a \emph{longest-match
  composition} operator. We will see some examples in Section \ref{subsec:packratparsing}.

@@

All the code in the paper and the case study is written in Scala, since its concise and elegant
syntax gives better presentation. Clearly
other languages that support some form of multiple inheritance
(including Java 8 with default methods~\cite{Goetz2012}) could in principle be
used.

### 2017-05-11 SLE_17/sections/ObjectAlgebras.tex
Format
Some changes.

The Expression Problem~\cite{wadler1998expression} illustrates the
difficulty of extending data structures or ASTs in two dimensions. For
traditional OO ASTs, adding new language constructs is easy, but
adding new operations is hard. In the last section, we have a simple
language containing literals, additions and variables. The
corresponding ASTs are built using classes with a pretty-printing
operation. Now suppose we want to collect free variables over the
ASTs. Using a traditional OO approach, one attempt would be extending
\inlinecode{Expr} with the new operation to obtain a new abstract type
for ASTs:

@@

\noindent Then all classes representing language constructs could be
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
even if that approach would work, the parsing code would not
be reusable. \lstinline{VarParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser have type \inlinecode{Parser[Expr]}, which indicates the result has type \inlinecode{Expr}, whereas the type should be \inlinecode{NewExpr} now.

Therefore, in order to generate new AST structures, we have to
substitute \lstinline{NewExpr}, \lstinline{NewLit},
\lstinline{NewAdd} and \lstinline{NewVar} for the old
\lstinline{Expr}, \lstinline{Lit}, \lstinline{Add} and \lstinline{Var}
correspondingly. Thus we have to either modify existing code or
use casts. For semantic modularity, we need a different approach for building ASTs.

@@

\lstinputlisting[linerange=5-8]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
We have two constructors \inlinecode{lit} and \inlinecode{add}, corresponding to literals and additions. The trait \inlinecode{Alg} is called an \textit{Object Algebra interface}. The trait is parametrized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

\paragraph{Adding New Operations}
To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing an expression:

@@

\lstinputlisting[linerange=33-35]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
In other words, existing code was reused and was not modified!

@@

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty-print the expression,
we pass an instance of \inlinecode{VarPrint} to \inlinecode{makeExp}:

\begin{lstlisting}
makeExp(new VarPrint {})
\end{lstlisting}
which results in \inlinecode{"(1 + x)"} as expected.

@@

\lstinputlisting[linerange=8-17]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser looks fine, but it is not extensible. For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunately, in this case, although \inlinecode{VarAlg} is a subtype
of \inlinecode{Alg}, the function type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{Alg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

@@

\noindent The type of the Object Algebra field \lstinline{alg} is first refined
to \lstinline{VarAlg[E]}, to allow calling the additional factory method
for variables. Note that, unlike the previous attempt, \lstinline{VarAlg[E]}
is a subtype of \lstinline{Alg[E]}, thus the type-refinement of the
field is allowed. Now, the code for parsing variables (\lstinline{pVar}) can
call \lstinline{alg.varE}. The following code illustrates how to use
the parser from a client's perspective:

\lstinputlisting[linerange=43-46]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

In the client code shown above, we first need to choose a concrete
Object Algebra to initialize the \lstinline{alg} field in \lstinline{VarOAParser}. Here we pick the
pretty-printing algebra \lstinline{VarPrint}, but any other Object
Algebra that implements \lstinline{VarAlg} would work.
With an instance of \lstinline{VarOAParser} in hand, we can call
\lstinline{pE} to obtain the parser to feed to the \lstinline{parse} method.
The result shows that such a pattern provides modular parsing.

Note that, similar to the approach in Section~\ref{sec:inheritance} independent extensibility is also supported via multiple trait inheritance.
Since independent extensibility is achieved using essentially the same technique as in Section~\ref{sec:inheritance} we omit that code here.

>>>>

The Expression Problem~\cite{wadler1998expression} illustrates the
difficulty of extending data structures or ASTs in two dimensions. For
traditional OO ASTs, adding new language constructs is easy, but
adding new operations is hard. In the last section, we have a simple
language containing literals, additions and variables. The
corresponding ASTs are built using classes with a pretty-printing
operation.
To modularly add an operation like collecting free variables, one attempt
would be extending
\inlinecode{Expr} with the new operation to obtain a new abstract type
for ASTs:

@@

\noindent Then all classes representing language constructs could be
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
even if that approach would work, the parsing code would not
be reusable. \lstinline{VarParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser have type \inlinecode{Parser[Expr]},
which generates ASTs that only support pretty-printing. Now we have to substitute \inlinecode{NewExpr} for \inlinecode{Expr} (the
same for \inlinecode{Lit}, \inlinecode{Add}, and so on). It requires either modification on existing code or casts.
The goal of semantic modularity motivates us to find a different approach for building ASTs.

@@

\lstinputlisting[linerange=5-8]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
Here \inlinecode{Alg} is called an \textit{Object Algebra interface}, parameterized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

\paragraph{Adding New Operations}
To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of pretty-printing:

@@

\lstinputlisting[linerange=33-35]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
Thus existing code was reused and was not modified!

@@

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty-print the expression, the code

\begin{lstlisting}
makeExp(new VarPrint {})
\end{lstlisting}
results in \inlinecode{"(1 + x)"} as expected.

@@

\lstinputlisting[linerange=8-17]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser looks fine, but it is not extensible. For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]}, which is a \emph{supertype} of the old type \inlinecode{Alg[E] =>} \inlinecode{Parser[E]}, since
the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This violates
overriding in Scala.

@@

\noindent The type of the Object Algebra field \lstinline{alg} is first refined
to \lstinline{VarAlg[E]}, to allow calling the additional factory method
for variables. Unlike the previous attempt, such a type-refinement is allowed.
Now, the code for parsing variables (\lstinline{pVar}) can
call \lstinline{alg.varE}. The following code illustrates how to use
the parser from a client's perspective:

\lstinputlisting[linerange=46-49]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

In the client code above,
we pick the pretty-printing algebra \lstinline{VarPrint} to initialize the \lstinline{alg} field, but any other Object
Algebra that implements \lstinline{VarAlg} would work.
With an instance of \lstinline{VarOAParser} in hand, we can call
\lstinline{pE} to obtain the parser to feed to the \lstinline{parse} method.
Such a pattern provides modular parsing as expected.

Note that, similar to the approach in Section~\ref{sec:inheritance}, independent extensibility is also supported via multiple trait inheritance.
Since independent extensibility is achieved using essentially the same technique as in Section~\ref{sec:inheritance}, we omit that code here.

### 2017-05-11 SLE_17/sections/PackratParsing.tex
Some changes.

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to compose them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, and so are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular, several parser combinators including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.

@@

Consider extending the language by adding subtractions to the grammar. If we still write its parser by directly following the grammar structure, the new left-recursive case will get the parser into an infinite loop. Namely, \inlinecode{parseExpr} and \inlinecode{parseSub} call each other and never stop.

@@

To solve this issue, we have to rewrite the grammar as below to eliminate left-recursion and build the parser based on this new grammar.

@@

In Parsec, for instance, the parser below will only parse the input \inlinecode{"1 + 2"} to \inlinecode{"1"}, as \inlinecode{parseInt} successfully parses \inlinecode{"1"}
and terminates parsing.

\begin{lstlisting}[language=PlainCode]
parser = parseInt <|> parseAdd
\end{lstlisting}

Using traditional alternative composition, when a preceding parser successfully parses a prefix of the input, it will finish parsing and return the result, in spite that subsequent parsers may be able to parse the whole input. In contrast with the parser above, \inlinecode{parser = parseAdd <|> parseInt} works as expected with the two cases swapped.

@@

\paragraph{Backtracking} The need for backtracking can also be problematic
in a modular setting. For example, consider a grammar that includes
``import'' statements including \inlinecode{import..from}.
Now we want to extend the grammar with an \inlinecode{import..as} case, as shown in the third line of grammar below.

\setlength{\grammarindent}{5em}
\begin{grammar}
<stmt> ::= `import' <ident> `from' <ident>
    \alt ...
    \alt `import' <ident> `as' <ident>
\end{grammar}

Since the \inlinecode{import..from} case shares a prefix with the new case \inlinecode{import..as}, when the former case fails, we must backtrack to the beginning. Take Parsec as an example, its choice
combinator \inlinecode{<|>} only tries the second alternative if the first fails
without any token consumption. An auxiliary function \inlinecode{try} is used for explicit backtracking.

\begin{lstlisting}[language=PlainCode]
oldParser = parseImpFrom <|> parseA <|> parseB <|> ...
newParser = try parseImpFrom <|> parseA <|> parseB <|> ... <|> parseImpAs
\end{lstlisting}

Given the full grammar, we can decide where to put \inlinecode{try} for backtracking. However, with modular parsing we are unable to have a global view of the full grammar. Hence the worst case should be considered that all alternatives may share common prefixes with future cases. In that case we need to backtrack for all the branches. To avoid failures in the future, we have to add \inlinecode{try} everywhere:

\begin{lstlisting}[language=PlainCode]
parser = try parseImpFrom <|> try parseA <|> try parseB <|> ... <|> try parseImpAs
\end{lstlisting}

\noindent However, this results in the worst-case exponential time
complexity in Parsec, because it does not provide optimizations
for backtracking.

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately, some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002} have been developed to address limitations of
simple parser combinators. Packrat parsers use memoization to record
the result of applying each parser at each position of the input, so
that repeated computation is eliminated. Moreover, Packrat parsing, in
theory, supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique for modular parsing.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
\cite{moors2008parser} which implements Packrat parsers.
The library provides a number of parser combinators, including the longest match alternative combinator.
%Below we present an example to illustrate Scala parsers.

@@

It is worth mentioning that the grammar is left-recursive in the addition case. Since left-recursion is well supported by Packrat parsers, we do not need extra code for it. Our parser also employs longest match composition by using the combinator \lstinline{|||} in the library. Furthermore, it does not suffer from the backtracking problem, as the memoization technique of Packrat parsing guarantees reasonable efficiency.

>>>>

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing.
One reason is that they are naturally modular for parser composition, providing full control to
the programmer. The other one is that they are implemented by a piece of code, so that it ensures
type safety in a statically typed programming language.
Unfortunately many parser combinators have important limitations.
In particular, several parser combinators including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.

@@

Consider extending the language by adding subtractions to the grammar.
If we directly extend the grammar, the new left-recursive case will cause an infinite loop.
Below, \inlinecode{parseExpr} and \inlinecode{parseSub} call each other and never stop.

@@

Instead, we have to rewrite the grammar as below to eliminate left-recursion.

@@

In Parsec, for instance, the parser \inlinecode{"parseInt <|> parseAdd"} will only parse the input \inlinecode{"1 + 2"} to \inlinecode{"1"}, as \inlinecode{parseInt} successfully parses \inlinecode{"1"}
and terminates parsing.

Traditional alternative composition will only find the first parser that succeeds on a prefix of the input,
in spite that subsequent parsers may parse the whole input. In contrast with the parser above, \inlinecode{"parseAdd <|> parseInt"} works as expected with the two cases swapped.

@@

\paragraph{Backtracking} The need for backtracking can also be problematic
in a modular setting. Consider a grammar that includes \inlinecode{"import..from"}.
It is extended with an \inlinecode{"import..as"} case:

\setlength{\grammarindent}{5em}
\begin{grammar}
<stmt> ::= `import' <ident> `from' <ident>
    \alt ...
    \alt `import' <ident> `as' <ident>
\end{grammar}
Since the two cases share a common prefix, when the former fails, we must backtrack to the beginning.
For example, the choice combinator in Parsec only tries the second alternative if the first fails
without any token consumption. We have to use \inlinecode{try} for explicit backtracking.

\begin{lstlisting}[language=PlainCode]
oldParser = parseImpFrom <|> ...
newParser = try parseImpFrom <|> ... <|> parseImpAs
\end{lstlisting}

Similarly, this violates a modular setting because we cannot have a global view of the full grammar.
Hence the worst case should be considered that all alternatives may share common prefixes with future cases. In that case we need to backtrack for all the branches. To avoid failures in the future, we have to add \inlinecode{try} everywhere. However this results in the worst-case exponential time
complexity.

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately, some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002} have been developed to address limitations of
simple parser combinators. Packrat parsing uses memoization to record
the result of applying each parser at each position of the input, so
that repeated computation is eliminated. Moreover, it supports both direct left-recursion
and (in theory) indirect left-recursion~\cite{warth2008}.
All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique for modular parsing.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
\cite{moors2008parser} for Packrat parsers.
The library provides a number of parser combinators, including the longest match alternative combinator.
%Below we present an example to illustrate Scala parsers.

@@

It is worth mentioning that the left-recursive grammar above is well supported without extra code.
 The longest match composition is also employed by using the combinator \lstinline{|||}. Furthermore, it does not suffer from the backtracking problem, as the memoization technique of Packrat parsing guarantees reasonable efficiency.
